---
title: "Psi1 snRNAseq"
subtitle: "Recluster Pass 1 Cell Types"
author: "Kennedi Todd"
date: "08/06/2025"
output:
  html_document:
    code_folding: hide
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

# Setup
## Set working directory
```{r working_directory}
knitr::opts_knit$set(root.dir = ".")
```

## Load libraries
```{r libraries, message=FALSE, warnings=FALSE}
library(dplyr)          # %>%
library(ggplot2)        # geom_vline()
library(Seurat)         # DimPlot()
library(SeuratWrappers) # RunPrestoAll()
```

## Variables and functions
```{r set_variables_and_thresholds}
# source thresholds and output paths
source("../../refs/thresholds_and_outs.R")
out <- paste0(out, "pass1/")
```

## Load data
```{r}
# save seurat object
mouse.annotated <- readRDS(
  file = paste0( "../../rObjects/pass1_annotated_seurat_obj.rds"),
)
```

## UMAP
```{r}
DimPlot(mouse.annotated, shuffle = TRUE)
```

# Neurons
## Subset
```{r}
# source thresholds and output paths
source("../../refs/thresholds_and_outs.R")
out <- paste0(out, "pass1/recluster_neurons/")

# subset neurons
neurons <- subset(mouse.annotated, 
                  individual_clusters %in% c("Neurons 1", "Neurons 2",
                                             "Neurons 3", "Neurons 4",
                                             "Neurons 5", "Neurons 6", 
                                             "Neurons 7", "Neurons 8",
                                             "Neurons 9", "Neurons 10",
                                             "Neurons 11", "Neurons 12", 
                                             "Neurons 13"))
table(neurons$individual_clusters)

# create new obj
options(Seurat.object.assay.calcn = TRUE)
counts <- GetAssayData(object = neurons, layer = "counts")
neurons <- CreateSeuratObject(counts)
neurons <- AddMetaData(neurons, 
                       metadata = mouse.annotated@meta.data[colnames(neurons), c(4:27)])

# cleanup data
remove(counts)
```

## Recluster
```{r recluster_neurons}
# split
neurons <- SplitObject(neurons, split.by = "sample_id")

# transform
options(future.globals.maxSize = 20000 * 1024^2)  # 20 GB
neurons <- lapply(neurons, SCTransform)

# get shared featured
shared_features <- SelectIntegrationFeatures(neurons)

# merge
neurons <- merge(neurons[[1]], y = neurons[-1])

# set default assay
DefaultAssay(neurons) <- "SCT"

# set variable features
VariableFeatures(neurons) <- shared_features

# PCA
neurons <- RunPCA(object = neurons, assay = "SCT")
DefaultAssay(neurons) <- "SCT"

# re-join layers
neurons[["RNA"]] <- JoinLayers(neurons[["RNA"]])

# elbow plot
ElbowPlot(neurons,
          ndims = 30,
          reduction = "pca") +
  geom_vline(xintercept = 20, linetype = "dashed", color = "red")

# run UMAP
neurons.unannotated <- RunUMAP(neurons,
                             dims = 1:20,
                             reduction = "pca",
                             n.components = 3)

# Determine the K-nearest neighbor graph
neurons.unannotated <- FindNeighbors(object = neurons.unannotated,
                                   assay = "SCT",
                                   reduction = "pca",
                                   dims = 1:20)

# Determine the clusters for various resolutions
neurons.unannotated <- FindClusters(object = neurons.unannotated,
                                  algorithm = 1,
                                  resolution = 0.5)

# Set default assay and normalize for visualization
DefaultAssay(neurons.unannotated) <- "RNA"
neurons.unannotated <- NormalizeData(neurons.unannotated)

# save
saveRDS(neurons.unannotated, 
        file = paste0( "../../rObjects/pass1_neurons_reclustered_unannotated.rds"),
        compress = FALSE)
```

## UMAP
```{r neurons_unannotated_umap}
Idents(neurons.unannotated) <- "seurat_clusters"

umap <- DimPlot(neurons.unannotated, shuffle = TRUE, label = TRUE)
umap
```

```{r save_neurons_unannotated_umap}
pdf(paste0(out, "unannotated_umap.pdf"), width = 6, height = 5)
umap
dev.off()
remove(umap)
```

## Markers
```{r neurons_unannotated_canonical_markers}
markers <- c(
  "Gja1", "P2ry12", "Flt1", "Ttr", "Col1a2", "Cfap44",      # non-neuronal
  "Spag16", "Mag", "Olig2", "Vtn", "Cd3e","Itgax","S100a9", # non-neuronal
  "Rbfox3", "Snap25",                                       # pan-neuronal
  "Gad1", "Gad2", "Sst", "Vip", "Pvalb", "Lamp5",           # GABAergic
  "Slc17a7", "Slc17a6", "Bcl11b", "Satb2", "Cux1",          # Glutamatergic
  "Cux2", "Foxp2", "Tle4", "Pcpr", "Rorb"                   # Glutamatergic
)

dot <- DotPlot(neurons.unannotated,
               features = markers,
               cols = "RdYlBu",
               assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dot
```

```{r save_neurons_unannotated_canonical_markers}
pdf(paste0(out, "unannotated_markers_dot_plot.pdf"), width = 12, height = 8)
dot
dev.off()
remove(dot)
```


## Assign
```{r assign_glut_gaba_identities}
new_cluster_names <- c(
  "0" = "Doublets",                   # Gad1, Gad2, Slc17a6
  "1" = "GABA Lamp5",                 # Gad1, Gad2, Lamp5
  "2" = "Glut",                       # Slc17a7, Rorb+++
  "3" = "Glut",                       # Slc17a7, Cux1+++, Cux2+++, Satb2+++
  "4" = "Glut",                       # Slc17a7,Foxp2+++, Tle4+++
  "5" = "Glut",                       # Slc17a7, Satb2+++
  "6" = "Glut",                       # Slc17a6, Rorb
  "7" = "Glut",                       # Slc17a7
  "8" = "GABA Pvalb",                 # Gad1, Gad2, Valb
  "9" = "Glut",                       # Slc17a7, Bcl11b
  "10" = "Doublets",                  # Gad1, Slc17a7
  "11" = "GABA Sst",                  # Gad1, Gad2, Sst
  "12" = "Glut",                      # Slc17a7, Tle4+++
  "13" = "GABA Lamp5",                # Gad1, Gad2, Lamp5+
  "14" = "GABA Vip",                  # Gad1, Gad2, Vip
  "15" = "Glut",                      # Slc17a7, Tle4+
  "16" = "Doublets",                  # Gad1, Gad2, Slc17a7
  "17" = "Glut",                      # Slc17a7, Slc17a6, Tle4+++
  "18" = "Glut",                      # Slc17a7, Bcl11b++, Satb2++
  "19" = "Glut",                      # Slc17a76, Slc17a7, Satb2+++
  "20" = "Glut",                      # Slc17a7, Slc17a6, Cux1+++, Cux2+++, Satb2+++
  "21" = "GABA Lamp5",                # Gad1, Gad1, Lamp5
  "22" = "Glut",                      # Slc17a7, Cux1, Cux2, Satb2
  "23" = "GABA",                      # Gad1, Gad2
  "24" = "Glut",                      # Slc17a7, Cux1, Cux2, Satb2
  "25" = "Doublets",                  # Gad1, Gad2, Slc17a7
  "26" = "Doublets",                  # Slc17a6, Cfap44
  "27" = "Glut",                      # Slc17a7, Cux1, Cux2, Satb2
  "28" = "Glut",                      # Slc17a7, Satb2
  "29" = "Glut",                      # Slc17a6, Rorb
  "30" = "Glut",                      # Sl17a7, Bcl11b, Rorb
  "31" = "GABA Pvalb",                # Gad1, Gad2, Pvalb
  "32" = "GABA Lamp5",                # Gad1, Gad2, Lamp5
  "33" = "GABA Pvalb",                # Gad1, Gad2, Pvalb
  "34" = "Glut",                      # Slc16a6, Cux1, Cux2
  "35" = "Doublets",                  # Gad2, Slc17a6
  "36" = "GABA",                      # Slc16a6, Gad1, Gad2
  "37" = "Glut",                      # Slc16a7, Satb2
  "38" = "Glut",                      # Slc16a6, Rorb
  "39" = "Doublets",                  # Gad1, Gad2, Slc17a7
  "40" = "GABA",                      # Gad1, Gad2
  "41" = "GABA Pvalb"                 # Gad1, Gad2, Pvalb
)

# set new idents
Idents(neurons.unannotated) <- "seurat_clusters"
neurons.annotated <- RenameIdents(neurons.unannotated, new_cluster_names)
neurons.annotated$neuron_class <- Idents(neurons.annotated)

# set level
neurons.annotated$neuron_class <- factor(neurons.annotated$neuron_class,
                                  levels = c("GABA",
                                             "GABA Lamp5",
                                             "GABA Pvalb",
                                             "GABA Sst",
                                             "GABA Vip",
                                             "Glut",
                                             "Doublets"))
Idents(neurons.annotated) <- "neuron_class"
```

```{r}
cluster_colors <- c("cyan","blue","yellow","purple","green","red","gray70")


# plot
u <- DimPlot(
  object = neurons.annotated,
  cols = cluster_colors,
  group.by = "neuron_class"
)
u
```

```{r astrocytes_barcodes_to_remove}
doublets <- subset(neurons.annotated, neuron_class == "Doublets")
barcodes <- colnames(doublets)
saveRDS(object = barcodes,
        file = "../../rObjects/pass1_recluster_neurons_barcodes_to_remove.rds",
        compress = FALSE)
remove(neurons, neurons.unannotated, neurons.annotated, doublets)
```

# Astrocytes
## Subset
```{r}
# source thresholds and output paths
source("../../refs/thresholds_and_outs.R")
out <- paste0(out, "pass1/recluster_astrocytes/")

# subset astrocytes
astrocytes <- subset(mouse.annotated, individual_clusters == "Astrocytes")

# create new obj
counts <- GetAssayData(object = astrocytes, layer = "counts")
astrocytes <- CreateSeuratObject(counts, meta.data = astrocytes@meta.data[,c(1,4:27)])

# cleanup data
remove(counts)
```

## Recluster
```{r recluster_astrocytes}
# split
astrocytes <- SplitObject(astrocytes, split.by = "sample_id")

# transform
options(future.globals.maxSize = 20000 * 1024^2)  # 20 GB
astrocytes <- lapply(astrocytes, SCTransform)

# get shared featured
shared_features <- SelectIntegrationFeatures(astrocytes)

# merge
astrocytes <- merge(astrocytes[[1]], y = astrocytes[-1])

# set default assay
DefaultAssay(astrocytes) <- "SCT"

# set variable features
VariableFeatures(astrocytes) <- shared_features

# PCA
astrocytes <- RunPCA(object = astrocytes, assay = "SCT")
DefaultAssay(astrocytes) <- "SCT"

# re-join layers
astrocytes[["RNA"]] <- JoinLayers(astrocytes[["RNA"]])

# elbow plot
ElbowPlot(astrocytes,
          ndims = 30,
          reduction = "pca") +
  geom_vline(xintercept = 11, linetype = "dashed", color = "red")

# run UMAP
astrocytes.unannotated <- RunUMAP(astrocytes,
                             dims = 1:10,
                             reduction = "pca",
                             n.components = 3)

# Determine the K-nearest neighbor graph
astrocytes.unannotated <- FindNeighbors(object = astrocytes.unannotated,
                                   assay = "SCT",
                                   reduction = "pca",
                                   dims = 1:10)

# Determine the clusters for various resolutions
astrocytes.unannotated <- FindClusters(object = astrocytes.unannotated,
                                  algorithm = 1,
                                  resolution = 0.1)

# Set default assay and normalize for visualization
DefaultAssay(astrocytes.unannotated) <- "RNA"
astrocytes.unannotated <- NormalizeData(astrocytes.unannotated)

# save
saveRDS(astrocytes.unannotated, 
        file = "../../rObjects/pass1_astrocytes_reclustered_unannotated.rds",
        compress = FALSE)
```

## UMAP
```{r astrocytes_unannotated_umap}
Idents(astrocytes.unannotated) <- "seurat_clusters"

umap <- DimPlot(astrocytes.unannotated, shuffle = TRUE, label = TRUE)
umap
```

```{r save_astrocytes_unannotated_umap}
pdf(paste0(out, "unannotated_umap.pdf"), width = 6, height = 5)
umap
dev.off()
remove(umap)
```

## Canonical markers
```{r astrocytes_unannotated_canonical_markers}
canonical_markers <- c(
  "Aldoc", "Aqp4", "Gja1", "Clu", "Apoe", "Slc1a2",   # Astrocytes
  "Aif1", "C1qa", "P2ry12", "Tmem119","Trem2","Mrc1", # Microglia
  "Ptprb", "Cldn5", "Flt1",                           # Endothelial
   "Ttr", "Folr1",                                    # Choroid
  "Col1a1", "Col1a2", "Dcn",                          # Fibroblast
  "Cfap44", "Spag16",                                 # Ependymal
  "Mog", "Opalin", "Mal", "Mag", "Mbp",               # Oligodendrocytes
  "Pdgfra", "Cspg4", "Olig1", "Olig2", "Tnr",         # OPCs
  "Syt1", "Snap25", "Gad1", "Meg3", "Igfbpl1","Sox4", # astrocytes
  "Pdgfrb", "Rgs5", "Acta2", "Vtn",                   # Mural
  "Cd3e","Itgax","S100a9"                             # Immune (T, DC, neutrophil)
)

dot <- DotPlot(astrocytes.unannotated,
               features = canonical_markers,
               cols = "RdYlBu",
               assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dot
```

```{r save_astrocytes_unannotated_canonical_markers}
pdf(paste0(out, "unannotated_canonical_markers_dot_plot.pdf"), width = 6, height = 9)
dot
dev.off()
remove(dot)
```

## Auto markers
```{r astrocytes_unannotated_auto_markers}
auto_markers <- SeuratWrappers::RunPrestoAll(
  object = astrocytes.unannotated,
  assay = "RNA",
  slot = "counts",
  only.pos = TRUE
)

top5 <- auto_markers %>%
  filter(p_val_adj < 0.01) %>%
  group_by(cluster) %>%
  slice_head(n = 5) %>%
  ungroup()

dot <- DotPlot(astrocytes.unannotated,
               features = unique(top5$gene),
               cols = "RdYlBu",
               assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dot
```

```{r save_astrocytes_unannotated_auto_markers}
pdf(paste0(out, "unannotated_auto_markers_dot_plot.pdf"), width = 7, height = 9)
dot
dev.off()
remove(dot)
```

```{r astrocytes_neuronal_marker_heatmap}
fplot <- FeaturePlot(
  object = astrocytes.unannotated,
  features = "Snap25"
)  + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
fplot
```

```{r astrocytes_barcodes_to_remove}
cluster3 <- subset(astrocytes.unannotated, seurat_clusters == "3")
barcodes <- colnames(cluster3)
saveRDS(object = barcodes,
        file = "../../rObjects/pass1_recluster_astrocytes_barcodes_to_remove.rds",
        compress = FALSE)
remove(astrocytes, astrocytes.unannotated, auto_markers, cluster3, fplot, top5)
```

# OPCs and Oligodendrocytes
## Subset
```{r subset_oligos}
# source thresholds and output paths
source("../../refs/thresholds_and_outs.R")
out <- paste0(out, "pass1/recluster_oligos_opcs/")

# subset oligos
oligos <- subset(mouse.annotated, 
                  individual_clusters %in% 
                   c("OPCs", "Differentiating OPCs", "Oligodendrocytes"))

# create new obj
counts <- GetAssayData(object = oligos, layer = "counts")
oligos <- CreateSeuratObject(counts, meta.data = oligos@meta.data[,c(1,4:27)])

# cleanup data
remove(counts)
```

## Recluster
```{r recluster_oligos}
# split
oligos <- SplitObject(oligos, split.by = "sample_id")

# transform
options(future.globals.maxSize = 20000 * 1024^2)  # 20 GB
oligos <- lapply(oligos, SCTransform)

# get shared featured
shared_features <- SelectIntegrationFeatures(oligos)

# merge
oligos <- merge(oligos[[1]], y = oligos[-1])

# set default assay
DefaultAssay(oligos) <- "SCT"

# set variable features
VariableFeatures(oligos) <- shared_features

# PCA
oligos <- RunPCA(object = oligos, assay = "SCT")
DefaultAssay(oligos) <- "SCT"

# re-join layers
oligos[["RNA"]] <- JoinLayers(oligos[["RNA"]])

# elbow plot
ElbowPlot(oligos,
          ndims = 30,
          reduction = "pca") +
  geom_vline(xintercept = 7, linetype = "dashed", color = "red")

# run UMAP
oligos.unannotated <- RunUMAP(oligos,
                             dims = 1:7,
                             reduction = "pca",
                             n.components = 3)

# Determine the K-nearest neighbor graph
oligos.unannotated <- FindNeighbors(object = oligos.unannotated,
                                   assay = "SCT",
                                   reduction = "pca",
                                   dims = 1:7)

# Determine the clusters for various resolutions
oligos.unannotated <- FindClusters(object = oligos.unannotated,
                                  algorithm = 1,
                                  resolution = 0.1)

# Set default assay and normalize for visualization
DefaultAssay(oligos.unannotated) <- "RNA"
oligos.unannotated <- NormalizeData(oligos.unannotated)

# save
saveRDS(oligos.unannotated, 
        file = paste0( "../../rObjects/", filtering_method, "_pass1_oligos_reclustered_unannotated.rds"),
        compress = FALSE)
```

## UMAP
```{r oligos_unannotated_umap}
Idents(oligos.unannotated) <- "seurat_clusters"

umap <- DimPlot(oligos.unannotated, shuffle = TRUE, label = TRUE)
umap
```

```{r save_oligos_unannotated_umap}
pdf(paste0(out, "unannotated_umap.pdf"), width = 6, height = 5)
umap
dev.off()
remove(umap)
```

## Canonical markers
```{r oligos_unannotated_canonical_markers}
canonical_markers <- c(
  "Aldoc", "Aqp4", "Gja1", "Clu", "Apoe", "Slc1a2",   # Astrocytes
  "Aif1", "C1qa", "P2ry12", "Tmem119","Trem2","Mrc1", # Microglia
  "Ptprb", "Cldn5", "Flt1",                           # Endothelial
   "Ttr", "Folr1",                                    # Choroid
  "Col1a1", "Col1a2", "Dcn",                          # Fibroblast
  "Cfap44", "Spag16",                                 # Ependymal
  "Mog", "Opalin", "Mal", "Mag", "Mbp",               # Oligodendrocytes
  "Pdgfra", "Cspg4", "Olig1", "Olig2", "Tnr",         # OPCs
  "Syt1", "Snap25", "Gad1", "Meg3", "Igfbpl1","Sox4", # oligos
  "Pdgfrb", "Rgs5", "Acta2", "Vtn",                   # Mural
  "Cd3e","Itgax","S100a9"                             # Immune (T, DC, neutrophil)
)

dot <- DotPlot(oligos.unannotated,
               features = canonical_markers,
               cols = "RdYlBu",
               assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dot
```

```{r save_oligos_unannotated_canonical_markers}
pdf(paste0(out, "unannotated_canonical_markers_dot_plot.pdf"), width = 6, height = 9)
dot
dev.off()
remove(dot)
```

## Auto markers
```{r oligos_unannotated_auto_markers}
auto_markers <- SeuratWrappers::RunPrestoAll(
  object = oligos.unannotated,
  assay = "RNA",
  slot = "counts",
  only.pos = TRUE
)

top5 <- auto_markers %>%
  filter(p_val_adj < 0.01) %>%
  group_by(cluster) %>%
  slice_head(n = 5) %>%
  ungroup()

dot <- DotPlot(oligos.unannotated,
               features = unique(top5$gene),
               cols = "RdYlBu",
               assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dot
```

```{r save_oligos_unannotated_auto_markers}
pdf(paste0(out, "unannotated_auto_markers_dot_plot.pdf"), width = 7, height = 9)
dot
dev.off()
remove(dot)
```

```{r oligos_neuronal_marker_heatmap}
fplot <- FeaturePlot(
  object = oligos.unannotated,
  features = "Cspg4"
)  + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
fplot
```

```{r oligos_barcodes_to_remove}
cluster5 <- subset(oligos.unannotated, seurat_clusters == "5")
barcodes <- colnames(cluster5)
saveRDS(object = barcodes,
        file = paste0("../../rObjects/", filtering_method, "_recluster_oligos_barcodes_to_remove.rds"))
remove(auto_markers, cluster5, fplot, oligos, oligos.unannotated, top5)
```


# Fibroblasts
## Subset
```{r subset_fibro}
# source thresholds and output paths
source("../../refs/thresholds_and_outs.R")
out <- paste0(out, "pass1/recluster_fibroblasts/")

# subset fibro
fibro <- subset(mouse.annotated, 
                  individual_clusters == "Fibroblasts")

# create new obj
counts <- GetAssayData(object = fibro, layer = "counts")
fibro <- CreateSeuratObject(counts, meta.data = fibro@meta.data[,c(1,4:27)])

# cleanup data
remove(counts)
```

## Recluster
```{r recluster_fibro}
# split
fibro <- SplitObject(fibro, split.by = "sample_id")

# transform
options(future.globals.maxSize = 20000 * 1024^2)  # 20 GB
fibro <- lapply(fibro, SCTransform)

# get shared featured
shared_features <- SelectIntegrationFeatures(fibro)

# merge
fibro <- merge(fibro[[1]], y = fibro[-1])

# set default assay
DefaultAssay(fibro) <- "SCT"

# set variable features
VariableFeatures(fibro) <- shared_features

# PCA
fibro <- RunPCA(object = fibro, assay = "SCT")
DefaultAssay(fibro) <- "SCT"

# re-join layers
fibro[["RNA"]] <- JoinLayers(fibro[["RNA"]])

# elbow plot
ElbowPlot(fibro,
          ndims = 30,
          reduction = "pca") +
  geom_vline(xintercept = 7, linetype = "dashed", color = "red")

# run UMAP
fibro.unannotated <- RunUMAP(fibro,
                             dims = 1:7,
                             reduction = "pca",
                             n.components = 3)

# Determine the K-nearest neighbor graph
fibro.unannotated <- FindNeighbors(object = fibro.unannotated,
                                   assay = "SCT",
                                   reduction = "pca",
                                   dims = 1:7)

# Determine the clusters for various resolutions
fibro.unannotated <- FindClusters(object = fibro.unannotated,
                                  algorithm = 1,
                                  resolution = 0.1)

# Set default assay and normalize for visualization
DefaultAssay(fibro.unannotated) <- "RNA"
fibro.unannotated <- NormalizeData(fibro.unannotated)

# save
saveRDS(fibro.unannotated, 
        file = paste0( "../../rObjects/", filtering_method, "_pass1_fibro_reclustered_unannotated.rds"),
        compress = FALSE)
```

## UMAP
```{r fibro_unannotated_umap}
Idents(fibro.unannotated) <- "seurat_clusters"

umap <- DimPlot(fibro.unannotated, shuffle = TRUE, label = TRUE)
umap
```

```{r save_fibro_unannotated_umap}
pdf(paste0(out, "unannotated_umap.pdf"), width = 6, height = 5)
umap
dev.off()
remove(umap)
```


# Choroid
## Subset
```{r subset_choroid}
# source thresholds and output paths
source("../../refs/thresholds_and_outs.R")
out <- paste0(out, "pass1/recluster_choroid/")

# subset choroid
choroid <- subset(mouse.annotated, 
                  individual_clusters == "Choroid")

# create new obj
counts <- GetAssayData(object = choroid, layer = "counts")
choroid <- CreateSeuratObject(counts, meta.data = choroid@meta.data[,c(1,4:27)])

# cleanup data
remove(counts)
```

## Recluster
```{r recluster_choroid}
# split
choroid <- SplitObject(choroid, split.by = "sample_id")

# transform
options(future.globals.maxSize = 20000 * 1024^2)  # 20 GB
choroid <- lapply(choroid, SCTransform)

# get shared featured
shared_features <- SelectIntegrationFeatures(choroid)

# merge
choroid <- merge(choroid[[1]], y = choroid[-1])

# set default assay
DefaultAssay(choroid) <- "SCT"

# set variable features
VariableFeatures(choroid) <- shared_features

# PCA
choroid <- RunPCA(object = choroid, assay = "SCT")
DefaultAssay(choroid) <- "SCT"

# re-join layers
choroid[["RNA"]] <- JoinLayers(choroid[["RNA"]])

# elbow plot
ElbowPlot(choroid,
          ndims = 30,
          reduction = "pca") +
  geom_vline(xintercept = 6, linetype = "dashed", color = "red")

# run UMAP
choroid.unannotated <- RunUMAP(choroid,
                             dims = 1:6,
                             reduction = "pca",
                             n.components = 3)

# Determine the K-nearest neighbor graph
choroid.unannotated <- FindNeighbors(object = choroid.unannotated,
                                   assay = "SCT",
                                   reduction = "pca",
                                   dims = 1:6)

# Determine the clusters for various resolutions
choroid.unannotated <- FindClusters(object = choroid.unannotated,
                                  algorithm = 1,
                                  resolution = 0.1)

# Set default assay and normalize for visualization
DefaultAssay(choroid.unannotated) <- "RNA"
choroid.unannotated <- NormalizeData(choroid.unannotated)

# save
saveRDS(choroid.unannotated, 
        file = paste0( "../../rObjects/", filtering_method, "_pass1_choroid_reclustered_unannotated.rds"),
        compress = FALSE)
```

## UMAP
```{r choroid_unannotated_umap}
Idents(choroid.unannotated) <- "seurat_clusters"

umap <- DimPlot(choroid.unannotated, shuffle = TRUE, label = TRUE)
umap
```

```{r save_choroid_unannotated_umap}
pdf(paste0(out, "unannotated_umap.pdf"), width = 6, height = 5)
umap
dev.off()
remove(umap)
```

## Canonical markers
```{r choroid_unannotated_canonical_markers}
canonical_markers <- c(
  "Aldoc", "Aqp4", "Gja1", "Clu", "Apoe", "Slc1a2",   # Astrocytes
  "Aif1", "C1qa", "P2ry12", "Tmem119","Trem2","Mrc1", # Microglia
  "Ptprb", "Cldn5", "Flt1",                           # Endothelial
   "Ttr", "Folr1",                                    # Choroid
  "Col1a1", "Col1a2", "Dcn",                          # Fibroblasts
  "Cfap44", "Spag16",                                 # Ependymal
  "Mog", "Opalin", "Mal", "Mag", "Mbp",               # Oligodendrocytes
  "Pdgfra", "Cspg4", "Olig1", "Olig2", "Tnr",         # OPCs
  "Syt1", "Snap25", "Gad1", "Meg3", "Igfbpl1","Sox4", # Neurons
  "Pdgfrb", "Rgs5", "Acta2", "Vtn",                   # Mural
  "Cd3e","Itgax","S100a9"                             # Immune (T, DC, neutrophil)
)

dot <- DotPlot(choroid.unannotated,
               features = canonical_markers,
               cols = "RdYlBu",
               assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dot
```

```{r save_choroid_unannotated_canonical_markers}
pdf(paste0(out, "unannotated_canonical_markers_dot_plot.pdf"), width = 6, height = 9)
dot
dev.off()
remove(dot)
```

## Auto markers
```{r choroid_unannotated_auto_markers}
auto_markers <- SeuratWrappers::RunPrestoAll(
  object = choroid.unannotated,
  assay = "RNA",
  slot = "counts",
  only.pos = TRUE
)

top5 <- auto_markers %>%
  filter(p_val_adj < 0.01) %>%
  group_by(cluster) %>%
  slice_head(n = 5) %>%
  ungroup()

dot <- DotPlot(choroid.unannotated,
               features = unique(top5$gene),
               cols = "RdYlBu",
               assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dot
```

```{r save_choroid_unannotated_auto_markers}
pdf(paste0(out, "unannotated_auto_markers_dot_plot.pdf"), width = 7, height = 9)
dot
dev.off()
remove(dot)
```

```{r choroid_neuronal_marker_heatmap}
fplot <- FeaturePlot(
  object = choroid.unannotated,
  features = "Cfap44"
)  + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
fplot
```

```{r choroid_barcodes_to_remove}
cluster0 <- subset(choroid.unannotated, seurat_clusters == "0")
barcodes <- colnames(cluster0)
saveRDS(object = barcodes,
        file = paste0("../../rObjects/", filtering_method, "_recluster_choroid_barcodes_to_remove.rds"))
remove(auto_markers, cluster0, fplot, choroid, choroid.unannotated, top5)
gc()
```


# Multiplets
## Subset
```{r subset_multiplets}
# source thresholds and output paths
source("../../refs/thresholds_and_outs.R")
out <- paste0(out, "pass1/recluster_multiplets/")

# subset multiplets
multiplets <- subset(mouse.annotated, 
                    individual_clusters %in% c("Multiplets 1", "Multiplets 2", 
                                               "Multiplets 3", "Multiplets 4", 
                                               "Multiplets 5", "Multiplets 6", 
                                               "Multiplets 7", "Multiplets 8", 
                                               "Multiplets 9"))

# create new obj
counts <- GetAssayData(object = multiplets, layer = "counts")
multiplets <- CreateSeuratObject(counts, meta.data = multiplets@meta.data[,c(1,4:27)])

# cleanup data
remove(counts)
```

## Recluster
```{r recluster_multiplets}
# split
multiplets <- SplitObject(multiplets, split.by = "sample_id")

# transform
options(future.globals.maxSize = 20000 * 1024^2)  # 20 GB
multiplets <- lapply(multiplets, SCTransform)

# get shared featured
shared_features <- SelectIntegrationFeatures(multiplets)

# merge
multiplets <- merge(multiplets[[1]], y = multiplets[-1])

# set default assay
DefaultAssay(multiplets) <- "SCT"

# set variable features
VariableFeatures(multiplets) <- shared_features

# PCA
multiplets <- RunPCA(object = multiplets, assay = "SCT")
DefaultAssay(multiplets) <- "SCT"

# re-join layers
multiplets[["RNA"]] <- JoinLayers(multiplets[["RNA"]])

# elbow plot
ElbowPlot(multiplets,
          ndims = 30,
          reduction = "pca") +
  geom_vline(xintercept = 11, linetype = "dashed", color = "red")

# run UMAP
multiplets.unannotated <- RunUMAP(multiplets,
                             dims = 1:11,
                             reduction = "pca",
                             n.components = 3)

# Determine the K-nearest neighbor graph
multiplets.unannotated <- FindNeighbors(object = multiplets.unannotated,
                                   assay = "SCT",
                                   reduction = "pca",
                                   dims = 1:11)

# Determine the clusters for various resolutions
multiplets.unannotated <- FindClusters(object = multiplets.unannotated,
                                  algorithm = 1,
                                  resolution = 0.2)

# Set default assay and normalize for visualization
DefaultAssay(multiplets.unannotated) <- "RNA"
multiplets.unannotated <- NormalizeData(multiplets.unannotated)

# save
saveRDS(multiplets.unannotated, 
        file = paste0( "../../rObjects/", filtering_method, "_pass1_multiplets_reclustered_unannotated.rds"),
        compress = FALSE)
```

## UMAP
```{r multiplets_unannotated_umap}
Idents(multiplets.unannotated) <- "seurat_clusters"

umap1 <- DimPlot(multiplets.unannotated, shuffle = TRUE, label = TRUE)
umap1

umap2 <- DimPlot(multiplets.unannotated, shuffle = TRUE, label = TRUE, dims = c(2,3))
umap2
```

```{r save_multiplets_unannotated_umap}
pdf(paste0(out, "unannotated_umap_dim1&2.pdf"), width = 6, height = 5)
umap1
dev.off()

pdf(paste0(out, "unannotated_umap_dim2&3.pdf"), width = 6, height = 5)
umap2
dev.off()

remove(umap1,umap2)
```

## Canonical markers
```{r multiplets_unannotated_canonical_markers}
canonical_markers <- c(
  "Aldoc", "Aqp4", "Gja1", "Clu", "Apoe", "Slc1a2",   # Astrocytes
  "Aif1", "C1qa", "P2ry12", "Tmem119","Trem2","Mrc1", # Microglia
  "Ptprb", "Cldn5", "Flt1",                           # Endothelial
   "Ttr", "Folr1",                                    # multiplets
  "Col1a1", "Col1a2", "Dcn",                          # Fibroblasts
  "Cfap44", "Spag16",                                 # Ependymal
  "Mog", "Opalin", "Mal", "Mag", "Mbp",               # Oligodendrocytes
  "Pdgfra", "Cspg4", "Olig1", "Olig2", "Tnr",         # OPCs
  "Syt1", "Snap25", "Gad1", "Meg3", "Igfbpl1","Sox4", # Neurons
  "Pdgfrb", "Rgs5", "Acta2", "Vtn",                   # Mural
  "Cd3e","Itgax","S100a9"                             # Immune (T, DC, neutrophil)
)

dot <- DotPlot(multiplets.unannotated,
               features = canonical_markers,
               cols = "RdYlBu",
               assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dot
```

```{r save_multiplets_unannotated_canonical_markers}
pdf(paste0(out, "unannotated_canonical_markers_dot_plot.pdf"), width = 8, height = 9)
dot
dev.off()
remove(dot)
```

## Auto markers
```{r multiplets_unannotated_auto_markers}
auto_markers <- SeuratWrappers::RunPrestoAll(
  object = multiplets.unannotated,
  assay = "RNA",
  slot = "counts",
  only.pos = TRUE
)

top3 <- auto_markers %>%
  filter(p_val_adj < 0.01) %>%
  group_by(cluster) %>%
  slice_head(n = 3) %>%
  ungroup()

dot <- DotPlot(multiplets.unannotated,
               features = unique(top5$gene),
               cols = "RdYlBu",
               assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
dot
```

```{r save_multiplets_unannotated_auto_markers}
pdf(paste0(out, "unannotated_auto_markers_dot_plot.pdf"), width = 8, height = 10)
dot
dev.off()
remove(dot)
```

```{r multiplets_neuronal_marker_heatmap}
fplot <- FeaturePlot(
  object = multiplets.unannotated,
  features = "Acta2",
  dims = c(2,3)
)  + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
fplot
```

```{r multiplets_barcodes_to_remove}
multiplets.unannotated$remove <- !multiplets.unannotated$seurat_clusters %in% c("8","13","16")
table(multiplets.unannotated$remove)
cluster <- subset(multiplets.unannotated, remove == TRUE)
barcodes <- colnames(cluster)
saveRDS(object = barcodes,
        file = paste0("../../rObjects/", filtering_method, "_recluster_multiplets_barcodes_to_remove.rds"))
remove(auto_markers, cluster, fplot, multiplets, multiplets.unannotated, top5)
gc()
```

