---
title: "Psilocybin Project 1"
subtitle: "Filtering and Normalization"
author: "Kennedi Todd"
date: "10/14/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

# Background
- bulk RNA sequencing of prefrontal cortex of 90 mice \
- 5 samples per group (treatment, dose, time point, sex) \
- high dose = 1 mg/kg, low dose = 0.25 mg/kg

# Setup
```{r setup, include=FALSE}
# Set working directory globally
knitr::opts_knit$set(root.dir = ".")
```

## Libraries
```{r libraries, message=FALSE, warning=FALSE}
library(caret)              # createDataPartition()
library(dotenv)             # load_dot_env()
library(dplyr)              # left_join()
library(edgeR)              # DGEList()
library(limma)              # plotMDS()
library(ggrepel)            # geom_text_repel()
library(ggplot2)            # ggplot()
library(glmnet)             # cv.glmnet()
library(gplots)             # heatmap.2()
library(grDevices)          # colorRampPalette()
library(gridExtra)          # grid.arrange()
library(knitr)              # opts_knit()
library(pheatmap)           # pheatmap()
library(philentropy)        # JSD()
library(reshape2)           # melt()
library(rtracklayer)        # import()
library(stringr)            # str_match()
library(sva)                # Combat_seq()
library(tidyverse)          # pivot_wider()
```

## User defined variables
```{r set_variables}
# Load environment variables
load_dot_env(file = "../refs/.env")  # Adjust the path to your .env file

# Output directories and file paths
out <- "../results/all_samples_sex_specific_combat_analysis/"
out2 <- "all_samples_sex_specific_combat_analysis"
ann <- Sys.getenv("MMUSCULUS_GTF")


# Colors and levels
treatment_order <- c("saline", "psilocybin")
treatment_colors <- c("gray60", "springgreen2")

sex_order <- c("M", "F")
sex_colors <- c("gold2", "purple")

group_order <- c("S.8h", "S.24h", "S.7d", 
                 "L.8h", "L.24h", "L.7d", 
                 "H.8h", "H.24h", "H.7d")
group_colors <- c("gray90", "gray60", "gray30", 
                  "lightblue", "cornflowerblue", "blue", 
                  "sienna1", "red2", "red4")
```

## Save functions
These functions with help simultaneously save plots as a png, pdf, and tiff 
file.
```{r save_functions}
saveToPDF <- function(...) {
    dev.copy(pdf, ...)
    dev.off()
}
```

# DGE object
## Read data
```{r read_data}
# Read counts data
counts <- read.delim("../featureCounts/gene_counts_matrix.tsv",
                     header = TRUE, row.names = 1, sep = "\t")

# Read metadata
meta <- read.delim("../refs/metadata.tsv", header = TRUE, sep = "\t") %>%
  arrange(factor(group, levels = group_order))
rownames(meta) <- meta$filename

# Rename counts columns
colnames(counts) <- gsub("_L[34]", "", colnames(counts))

# Ensure counts columns match metadata rows
counts <- counts[, meta$filename]
stopifnot(all.equal(colnames(counts), meta$filename))

# Use sample_id for row/column names
rownames(meta) <- meta$sample_id
colnames(counts) <- meta$sample_id

# Read or create annotation file
annotation_rds <- "../rObjects/annotation.rds"
if (file.exists(annotation_rds)) {
  genes <- readRDS(annotation_rds)
} else {
  genes <- rtracklayer::import(ann) %>%
    as.data.frame() %>%
    filter(type == "gene")
  saveRDS(genes, annotation_rds)
}

# Check rownames match
stopifnot(all.equal(rownames(counts), genes$gene_id))

# Read exonic length information and add to genes
exonic_length <- read.delim("../featureCounts/summed_exons.tsv", header = TRUE, sep = "\t")
genes <- left_join(genes, exonic_length, by = "gene_id")
rownames(counts) <- make.unique(genes$gene_name)
genes$gene_name_unique <- rownames(counts)
```

## Create object
```{r DGE_object}
# Create DGE object
dge <- DGEList(counts = counts, genes = genes, samples = meta)

# Set levels for factors
dge$samples <- dge$samples %>%
  mutate(treatment = factor(treatment, levels = treatment_order),
         group = factor(group, levels = group_order),
         sex = factor(sex, levels = sex_order),
         lane = factor(lane))

# Cleanup
rm(counts, meta, genes, exonic_length)
```

## Remove genes
```{r remove_MT_genes}
# Remove mitochondrial genes (chrM)
dge <- dge[dge$genes$seqnames != "chrM", , keep.lib.sizes = FALSE]

# Keep only protein-coding genes
dge <- dge[dge$genes$gene_type == "protein_coding", , keep.lib.sizes = FALSE]

# Calculate Hbb-bs log2 CPM and add to metadata
dge$samples$hbb_log2cpm <- as.numeric(cpm(dge, log = TRUE)["Hbb-bs", ])
```

## Remove samples
```{r}
# if subset by time point
if (grepl("8_hours", out2)) {
  print("8 hours only")
  dge <- dge[,dge$samples$timepoint == "8 hour"]
  group_order <- c("S.8h","L.8h","H.8h")
  group_colors <- c("gray60", "cornflowerblue", "red2")
} else if (grepl("24_hours", out2)) {
  print("24 hours only")
  dge <- dge[,dge$samples$timepoint == "24 hour"]
  group_order <- c("S.24h","L.24h","H.24h")
  group_colors <- c("gray60", "cornflowerblue", "red2")
} else if (grepl("7_days", out2)) {
  print("7 days only")
  dge <- dge[,dge$samples$timepoint == "7 day"]
  group_order <- c("S.7d","L.7d","H.7d")  
  group_colors <- c("gray60", "cornflowerblue", "red2")
}

head(colnames(dge))
dim(dge)
```

## ComBat
```{r}
# if a ComBat analysis
if (grepl("combat", out2)) {
  # extract meta
  meta <- dge$samples
  meta[c(1,5,6,7,8,9,11)] <- lapply(meta[c(1,5,6,7,8,9,11)], factor)
  # create model
  model <- model.matrix(~ dose + sex, data = meta)
  # adjust counts for batch correction
  adjusted_counts <- ComBat_seq(counts = dge$counts,
                                batch = meta$lane,
                                group = NULL, 
                                covar_mod = model)
  # update DGEList
  dge <- DGEList(counts = adjusted_counts, 
                 genes = dge$genes, 
                 samples = meta)
}
dim(dge)
```

# QC: Raw
## Bargraph
```{r cpm_bar_graph}
# user input
goi <- c("Eif2s3y","Uty","Kdm5d","Ddx3y","Eif2s3x","Ddx3x","Kdm6a","Kdm5c",
         "Med12","Hbb-bs")

# plot and save
for (i in 1:length(goi)) {

  # extract data
  df <- as.data.frame(cpm(dge$counts))
  df <- df[goi[i],]
  df <- reshape2::melt(df)
  colnames(df) <- c("sample","CPM")
  df$sample <- as.character(df$sample)
  df$sex <- dge$samples$sex
  df <- df[order(df$sex),]
  
  # plot
  p <- ggplot(data = df, mapping = aes(x = sample, y = CPM, fill = sex)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = paste("Raw:", goi[i], "CPM"), x = "Sample", y = "CPM") +
    scale_x_discrete(limits = df$sample)
  print(p)
  
  # save
  path <- paste0(out, "prefiltering_QC/", goi[i], "_bar_graph_raw")
  saveToPDF(paste0(path, ".pdf"), width = 12, height = 4)
  dev.off()
}
remove(df,p)
```

## JSD heatmap
This portion won't display in the R Markdown pdf; the margins are too large.
The pdf and png file can only be saved one at a time.
```{r JSD_heatmap}
# find JSD
data <- JSD(t(cpm(dge$counts)), est.prob = "empirical")
names <- dge$samples$sample_id
colnames(data) <- names
rownames(data) <- names
round.data <- round(data, digits = 4) # round 2 decimal places

# save
path <- paste0(out,"prefiltering_QC/JSD_heatmap.pdf")
pdf(path, width = 15, height = 15)

# set heatmap colors and names
meta <- dge$samples[,c("treatment","sex","timepoint")]
rownames(meta) <- names
paletteLength <- 100
myColor <- colorRampPalette(c("blue", "skyblue", "white"))(paletteLength)
ann_colors = list(treatment = c(saline="lightgray", 
                                psilocybin="springgreen2"),
                  sex = c(M="gold", 
                          F="purple"),
                  timepoint = c(`8 hour`="tomato",
                                `24 hour`="cornflowerblue",
                                `7 day`="tan"))

pheatmap::pheatmap(round.data,
                   treeheight_row = 0,
                   treeheight_col = 0,
                   annotation_row = meta,
                   annotation_col = meta,
                   annotation_colors = ann_colors,
                   color = myColor,
                   #display_numbers = TRUE,
                   #fontsize_number = 6,
                   number_color = "black")
remove(round.data,ann_colors)
```

## MDS plots
- MDS = multidimensional scaling \
- limma::plotMDS() will produce a PCoA or PCA plot \
- By default, gene.selection = "pairwise" \
- If gene.selection = "pairwise" and top is < nrow(x) then a PCoA plot is produced \
- Distances on the plot represent the leading log2 fold-changes \
- The leading log fold-change between a pair of samples is the root-mean-square average of the top largest log2 fold-changes between those two samples \
### Sex
```{r mds_colored_by_sex}
# set colors and get data
data <- cpm(dge$counts, log = TRUE)
names <- dge$samples$sample_id

# loop through dimensions and top genes
dims <- list(c(1,2),c(2,3),c(3,4))
topGenes <- c(100,500)

for (dim in dims) {
  
  for (numGenes in topGenes) {
    
    # Create MDS plot  
    mds <- plotMDS(data, 
                   labels = names, 
                   top = numGenes,
                   plot = FALSE,
                   dim.plot = dim)

    # Prepare data for plotting
    toplot <- data.frame(DimX = mds$x, 
                         DimY = mds$y,
                         Sample = rownames(mds$distance.matrix.squared),
                         sex = dge$samples$sex)
    
    # Set axes labels
    x.lab <- paste0(mds$axislabel," ",mds$dim.plot[1]," (",
                    round(mds$var.explained[mds$dim.plot[1]] * 100, digits = 2),
                    "%)")
    y.lab <- paste0(mds$axislabel," ",mds$dim.plot[2]," (",
                    round(mds$var.explained[mds$dim.plot[2]] * 100, digits = 2),
                    "%)")
    
    # Plot using ggplot
    p <- ggplot(toplot, aes(x = DimX, y = DimY, color = sex, label = Sample)) + 
      geom_point(size = 3) +
      scale_color_manual(values = sex_colors) +
      theme_bw() +
      geom_text(aes(label = Sample), size=4, vjust = -1) +
      xlab(x.lab) + ylab(y.lab) + 
      ggtitle(paste0("MDS Raw: Top ", numGenes, ", Log2(CPM)"))
    print(p)
    
    # save
    dimTitle <- as.character(dim)
    path <- paste0(out,"prefiltering_QC/MDS_top_",
                   numGenes,"_genes/MDS_raw_top_", numGenes, "_",
                   "dim", as.character(dim)[1], "&", as.character(dim[2]), 
                   "_colored_by_sex")
    saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
  }
}

remove(data)
```

### Group
```{r mds_colored_by_group}
# set colors and get data
data <- cpm(dge$counts, log = TRUE)
names <- dge$samples$informative_name

# loop through dimensions and top genes
dims <- list(c(1,2),c(2,3),c(3,4))
topGenes <- c(100,500)

for (dim in dims) {
  
  for (numGenes in topGenes) {
      
    mds <- plotMDS(data, 
                   labels = names, 
                   top = numGenes,
                   plot = FALSE,
                   dim.plot = dim)

    toplot <- data.frame(DimX = mds$x, 
                         DimY = mds$y,
                         Sample = rownames(mds$distance.matrix.squared),
                         group = dge$samples$group)
    
    x.lab <- paste0(mds$axislabel," ",mds$dim.plot[1]," (",
                    round(mds$var.explained[mds$dim.plot[1]] * 100, digits = 2),
                    "%)")
    y.lab <- paste0(mds$axislabel," ",mds$dim.plot[2]," (",
                    round(mds$var.explained[mds$dim.plot[2]] * 100, digits = 2),
                    "%)")
    
    p <- ggplot(toplot, aes(x = DimX, y = DimY, color = group, label = Sample)) + 
      geom_point(size = 3) +
      scale_color_manual(values = group_colors) +
      theme_bw() +
      geom_text(aes(label = Sample), size=4, vjust = -1) +
      xlab(x.lab) + ylab(y.lab) + 
      ggtitle(paste0("MDS Raw: Top ", numGenes, ", Log2(CPM)"))
    print(p)
    
    # save
    dimTitle <- as.character(dim)
    path <- paste0(out,"prefiltering_QC/MDS_top_",
                   numGenes,"_genes/MDS_raw_top_", numGenes, "_",
                   "dim", as.character(dim)[1], "&", as.character(dim[2]), 
                   "_colored_by_group")
    saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
  }
}

remove(data)
```

### Treatment
```{r mds_colored_by_treatment}
# set colors and get data
data <- cpm(dge$counts, log = TRUE)
names <- dge$samples$sample_id

# loop through dimensions and top genes
dims <- list(c(1,2),c(2,3),c(3,4))
topGenes <- c(100,500)

for (dim in dims) {
  
  for (numGenes in topGenes) {
      
    mds <- plotMDS(data, 
                   labels = names, 
                   top = numGenes,
                   plot = FALSE,
                   dim.plot = dim)

    toplot <- data.frame(DimX = mds$x, 
                         DimY = mds$y,
                         Sample = rownames(mds$distance.matrix.squared),
                         treatment = dge$samples$treatment)
    
    x.lab <- paste0(mds$axislabel," ",mds$dim.plot[1]," (",
                    round(mds$var.explained[mds$dim.plot[1]] * 100, digits = 2),
                    "%)")
    y.lab <- paste0(mds$axislabel," ",mds$dim.plot[2]," (",
                    round(mds$var.explained[mds$dim.plot[2]] * 100, digits = 2),
                    "%)")
    
    p <- ggplot(toplot, aes(x = DimX, y = DimY, color = treatment, label = Sample)) + 
      geom_point(size = 3) +
      scale_color_manual(values = treatment_colors) +
      theme_bw() +
      geom_text(aes(label = Sample), size=4, vjust = -1) +
      xlab(x.lab) + ylab(y.lab) + 
      ggtitle(paste0("MDS Raw: Top ", numGenes, ", Log2(CPM)"))
    print(p)
    
    # save
    dimTitle <- as.character(dim)
    path <- paste0(out,"prefiltering_QC/MDS_top_",
                   numGenes,"_genes/MDS_raw_top_", numGenes, "_",
                   "dim", as.character(dim)[1], "&", as.character(dim[2]), 
                   "_colored_by_treatment")
    saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
  }
}

remove(data)
```

### Time point
```{r mds_colored_by_timepoint}
# set colors and get data
data <- cpm(dge$counts, log = TRUE)
names <- dge$samples$sample_id

# loop through dimensions and top genes
dims <- list(c(1,2),c(2,3),c(3,4))
topGenes <- c(100,500)

for (dim in dims) {
  
  for (numGenes in topGenes) {
      
    mds <- plotMDS(data, 
                   labels = names, 
                   top = numGenes,
                   plot = FALSE,
                   dim.plot = dim)

    toplot <- data.frame(DimX = mds$x, 
                         DimY = mds$y,
                         Sample = rownames(mds$distance.matrix.squared),
                         timepoint = dge$samples$timepoint)
    
    x.lab <- paste0(mds$axislabel," ",mds$dim.plot[1]," (",
                    round(mds$var.explained[mds$dim.plot[1]] * 100, digits = 2),
                    "%)")
    y.lab <- paste0(mds$axislabel," ",mds$dim.plot[2]," (",
                    round(mds$var.explained[mds$dim.plot[2]] * 100, digits = 2),
                    "%)")
    
    p <- ggplot(toplot, aes(x = DimX, y = DimY, color = timepoint, label = Sample)) + 
      geom_point(size = 3) +
      theme_bw() +
      geom_text(aes(label = Sample), size=4, vjust = -1) +
      xlab(x.lab) + ylab(y.lab) + 
      ggtitle(paste0("MDS Raw: Top ", numGenes, ", Log2(CPM)"))
    print(p)
    
    # save
    dimTitle <- as.character(dim)
    path <- paste0(out,"prefiltering_QC/MDS_top_",
                   numGenes,"_genes/MDS_raw_top_", numGenes, "_",
                   "dim", as.character(dim)[1], "&", as.character(dim[2]), 
                   "_colored_by_timepoint")
    saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
  }
}

remove(data)
```

### Dose
```{r mds_colored_by_dose}
# set colors and get data
dge$samples$dose <- factor(dge$samples$dose)
data <- cpm(dge$counts, log = TRUE)
names <- dge$samples$sample_id

# loop through dimensions and top genes
dims <- list(c(1,2),c(2,3),c(3,4))
topGenes <- c(100,500)

for (dim in dims) {
  
  for (numGenes in topGenes) {
      
    mds <- plotMDS(data, 
                   labels = names, 
                   top = numGenes,
                   plot = FALSE,
                   dim.plot = dim)

    toplot <- data.frame(DimX = mds$x, 
                         DimY = mds$y,
                         Sample = rownames(mds$distance.matrix.squared),
                         Dose = dge$samples$dose)
    
    x.lab <- paste0(mds$axislabel," ",mds$dim.plot[1]," (",
                    round(mds$var.explained[mds$dim.plot[1]] * 100, digits = 2),
                    "%)")
    y.lab <- paste0(mds$axislabel," ",mds$dim.plot[2]," (",
                    round(mds$var.explained[mds$dim.plot[2]] * 100, digits = 2),
                    "%)")
    
    p <- ggplot(toplot, aes(x = DimX, y = DimY, color = Dose, label = Sample)) + 
      geom_point(size = 3) +
      scale_color_manual(values = c("red","green3","blue")) +
      theme_bw() +
      geom_text(aes(label = Sample), size=4, vjust = -1) +
      xlab(x.lab) + ylab(y.lab) + 
      ggtitle(paste0("MDS Raw: Top ", numGenes, ", Log2(CPM)"))
    print(p)
    
    # save
    dimTitle <- as.character(dim)
    path <- paste0(out,"prefiltering_QC/MDS_top_",
                   numGenes,"_genes/MDS_raw_top_", numGenes, "_",
                   "dim", as.character(dim)[1], "&", as.character(dim[2]), 
                   "_colored_by_dose")
    saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
  }
}

remove(data,dims)
```

### RIN
```{r MDS_colored_by_rin}
# set colors and get data
dge$samples$RNA_extraction_batch <- factor(dge$samples$RNA_extraction_batch)
data <- cpm(dge$counts, log = TRUE)
names <- dge$samples$sample_id

# loop through dimensions and top genes
dims <- list(c(1,2),c(2,3),c(3,4))
topGenes <- c(100,500)

for (dim in dims) {
  
  for (numGenes in topGenes) {
      
    mds <- plotMDS(data, 
                   labels = names, 
                   top = numGenes,
                   plot = FALSE,
                   dim.plot = dim)

    toplot <- data.frame(DimX = mds$x, 
                         DimY = mds$y,
                         Sample = rownames(mds$distance.matrix.squared),
                         RIN = dge$samples$RIN)
    
    x.lab <- paste0(mds$axislabel," ",mds$dim.plot[1]," (",
                    round(mds$var.explained[mds$dim.plot[1]] * 100, digits = 2),
                    "%)")
    y.lab <- paste0(mds$axislabel," ",mds$dim.plot[2]," (",
                    round(mds$var.explained[mds$dim.plot[2]] * 100, digits = 2),
                    "%)")
    
    p <- ggplot(toplot, aes(x = DimX, y = DimY, color = RIN, label = Sample)) + 
      geom_point(size = 3) +
      theme_bw() +
      geom_text(aes(label = Sample), size=4, vjust = -1) +
      xlab(x.lab) + ylab(y.lab) + 
      scale_color_gradientn(colors = c("red","yellow","lightblue","blue")) +
      ggtitle(paste0("MDS Raw: Top ", numGenes, ", Log2(CPM)"))
    print(p)
    
    # save
    dimTitle <- as.character(dim)
    path <- paste0(out,"prefiltering_QC/MDS_top_",
                   numGenes,"_genes/MDS_raw_top_", numGenes, "_",
                   "dim", as.character(dim)[1], "&", as.character(dim[2]), 
                   "_colored_by_RIN")
    saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
  }
}

remove(data,dims,p,toplot)
```

### RNA extraction
```{r mds_colored_by_dose}
# set colors and get data
dge$samples$RNA_extraction_batch <- factor(dge$samples$RNA_extraction_batch)
data <- cpm(dge$counts, log = TRUE)
names <- dge$samples$sample_id

# loop through dimensions and top genes
dims <- list(c(1,2),c(2,3),c(3,4))
topGenes <- c(100,500)

for (dim in dims) {
  
  for (numGenes in topGenes) {
      
    mds <- plotMDS(data, 
                   labels = names, 
                   top = numGenes,
                   plot = FALSE,
                   dim.plot = dim)

    toplot <- data.frame(DimX = mds$x, 
                         DimY = mds$y,
                         Sample = rownames(mds$distance.matrix.squared),
                         RNA_extraction_batch = dge$samples$RNA_extraction_batch)
    
    x.lab <- paste0(mds$axislabel," ",mds$dim.plot[1]," (",
                    round(mds$var.explained[mds$dim.plot[1]] * 100, digits = 2),
                    "%)")
    y.lab <- paste0(mds$axislabel," ",mds$dim.plot[2]," (",
                    round(mds$var.explained[mds$dim.plot[2]] * 100, digits = 2),
                    "%)")
    
    p <- ggplot(toplot, aes(x = DimX, y = DimY, color = RNA_extraction_batch, label = Sample)) + 
      geom_point(size = 3) +
      #scale_color_manual(values = c("red","green3","blue","purple")) +
      theme_bw() +
      geom_text(aes(label = Sample), size=4, vjust = -1) +
      xlab(x.lab) + ylab(y.lab) + 
      ggtitle(paste0("MDS Raw: Top ", numGenes, ", Log2(CPM)"))
    print(p)
    
    # save
    dimTitle <- as.character(dim)
    path <- paste0(out,"prefiltering_QC/MDS_top_",
                   numGenes,"_genes/MDS_raw_top_", numGenes, "_",
                   "dim", as.character(dim)[1], "&", as.character(dim[2]), 
                   "_colored_by_RNA_extraction")
    saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
  }
}

remove(data,dims)
```

## MDS variance
```{r mds_variance_explained}
# Set colors and get data
data <- cpm(dge$counts, log = TRUE)
names <- dge$samples$sample_id

# Define your variables once
variables <- list(
  sex = factor(dge$samples$sex),
  group = factor(dge$samples$group),
  treatment = factor(dge$samples$treatment),
  dose = factor(dge$samples$dose),
  RIN = as.numeric(dge$samples$RIN),
  lane = factor(dge$samples$lane),
  timepoint = factor(dge$samples$timepoint),
  RNA_extraction_batch = factor(dge$samples$RNA_extraction_batch)
)

# List of dimensions to iterate over
dimensions <- c(1, 2, 3, 4)

# Initialize an empty matrix to store R-squared values
r_squared_matrix <- matrix(NA, nrow = length(variables), ncol = length(dimensions))
rownames(r_squared_matrix) <- names(variables)
colnames(r_squared_matrix) <- paste0("Dim", dimensions)

# Prepare the main data frame once with MDS dimensions and sample-level variables
mds_data <- data.frame(
  Sample = rownames(dge$samples),
  variables
)

# First loop: iterate over variables
for (var in names(variables)) {
  
  # Second loop: iterate over MDS dimensions (one dimension at a time)
  for (dim in dimensions) {
    
    # Perform MDS analysis for the top 500 genes and the specific dimension pair
    mds <- plotMDS(data, 
                   labels = names, 
                   top = 500, 
                   plot = FALSE, 
                   dim.plot = c(dim, dim + 1))  # Ensure two dimensions
    
    # Add MDS Dimension X (Dim1, Dim2, etc.) to the data frame
    mds_data$DimX <- mds$x
    
    # Check if the variable is categorical or continuous
    if (is.factor(mds_data[[var]]) || is.character(mds_data[[var]])) {
      # Convert categorical variables to numeric factors
      mds_data[[paste0(var, "_numeric")]] <- as.numeric(as.factor(mds_data[[var]]))
    } else {
      # For continuous variables, just use the variable directly
      mds_data[[paste0(var, "_numeric")]] <- mds_data[[var]]
    }
    
    # Fit a linear model: MDS Dimension X ~ variable
    lm_model_X <- lm(DimX ~ mds_data[[paste0(var, "_numeric")]], data = mds_data)
    
    # Get the R-squared value for the current dimension and round it
    r_squared <- round(summary(lm_model_X)$r.squared, 3)
    
    # Store the R-squared value in the matrix
    r_squared_matrix[var, paste0("Dim", dim)] <- r_squared
  }
}

# Cleanup intermediate variables, keeping only the final R-squared matrix
rm(data, names, variables, dimensions, var, dim, mds, mds_data, lm_model_X, r_squared)

# Display the R-squared matrix
print(r_squared_matrix)

# Save
write.table(x = r_squared_matrix,
            file = paste0(out, "prefiltering_QC/variance_explained_by_MDS_dims.tsv"),
            sep = "\t",
            quote = FALSE)

# cleanup
remove(r_squared_matrix)
```

## CPM, FPKM, TPM
```{r prefiltering_cpm_fpkm_tpm}
# CPM, Counts Per Million
# normalize by sequencing depth
cpm <- cpm(dge)
path <- paste0(out, "counts/CPM_prefiltering.tsv")
write.table(cpm, path, quote = FALSE, sep = "\t")

# FPKM, Fragments Per Kilobase Million
# normalize by sequencing depth and then length
exonic.length <- dge$genes$exonic_length
exonic.length.kb <- exonic.length / 1000
fpkm <- cpm / exonic.length.kb
fpkm <- sweep(cpm, 1, exonic.length.kb, FUN = '/') # 1 = row
path <- paste0(out, "counts/FPKM_prefiltering.tsv")
write.table(fpkm, path, quote = FALSE, sep = "\t")

# TPM, Transcripts Per Kilobase Million
# normalize by length and then sequencing depth
counts <- dge$counts
fpk <- counts / exonic.length.kb
sum.col <- colSums(fpk) / 1000000
tpm <- sweep(fpk, 2, sum.col, FUN = '/') # 2 = column
path <- paste0(out, "counts/TPM_prefiltering.tsv")
write.table(tpm, path, quote = FALSE, sep = "\t")

remove(counts,cpm,fpk,fpkm,tpm)
```

# Filtering and normalization
## Lowly expressed genes
```{r filter_genes_by_expr}
if (grepl("both_sexes", out2)) {
  
  print("both sexes analysis")
  
  # get the number of genes we are starting with
  start <- dim(dge)[1]
  
  # get cpm dataframe
  cpm <- as.data.frame(cpm(dge))
  
  # get number of samples per group (i.e. S.24h)
  meta <- dge$samples[,c("sample_id","group")] |> 
    group_by(group) |> 
    mutate(group_size = n())
  
  # find average CPM per group
  avgPerGroup <- cpm |>
    mutate(gene = row.names(cpm)) |>
    pivot_longer(-gene, names_to = "sample_id") |>
    left_join(meta, by = "sample_id") |>
    summarize(mean = mean(value), .by = c(gene, group)) |>
    pivot_wider(names_from = group, values_from = mean) |>
    column_to_rownames(var = "gene") |>
    round(digits = 4)
  
  # save avg CPM per group
  write.table(x = avgPerGroup, 
              file = paste0(out,"counts/avgCPM_per_group_before_filtering.tsv"),
              quote = FALSE,
              sep = "\t")
  
  # all samples in a group (i.e. S.8h) must have CPM >= 1
  # create df with summary of if group has all samples >= 1  
  filterTable <- cpm |>
    mutate(across(everything(), ~ . >= 1)) |>
    mutate(gene = rownames(cpm)) |>
    pivot_longer(-gene, names_to = "sample_id") |>
    left_join(meta, by = "sample_id") |>
    summarize(thresh = all(value), .by = c(gene, group)) |>
    pivot_wider(names_from = group, values_from = thresh) |>
    column_to_rownames(var = "gene") 
  
  # add a 'pass' column that checks if any group meets the threshold
  filterTable$pass <- rowSums(filterTable) > 0
  
  # compare row names with 'dge'
  all.equal(rownames(filterTable),rownames(dge))
  
  # filter
  dge.filtered <- dge[filterTable$pass,,keep.lib.sizes = FALSE]
  
  # save
  avgPerGroupFiltered <- avgPerGroup[filterTable$pass,]
  write.table(x = avgPerGroupFiltered, 
              file = paste0(out, "counts/avgCPM_per_group_after_filtering.tsv"),
              quote = FALSE,
              sep = "\t")
  
  # see number of genes leftover
  end <- dim(dge.filtered)[1]
  
  # print message
  print(paste(start - end, "genes removed and", end, "genes leftover."))
  
  # cleanup
  remove(meta,avgPerGroup,cpm,start,end,filterTable)
  
} else if (grepl("sex_specific", out2)) {
  
  print("sex specific filtering")
    
  # get the number of genes we are starting with
  start <- dim(dge)[1]
  
  # get cpm dataframe
  cpm <- as.data.frame(cpm(dge))
  
  # get number of samples per group2 (i.e. S.24h)
  meta <- dge$samples[,c("sample_id","group2")] |> 
    group_by(group2) |> 
    mutate(group_size = n())
  
  # find average CPM per group2
  avgPerGroup <- cpm |>
    mutate(gene = row.names(cpm)) |>
    pivot_longer(-gene, names_to = "sample_id") |>
    left_join(meta, by = "sample_id") |>
    summarize(mean = mean(value), .by = c(gene, group2)) |>
    pivot_wider(names_from = group2, values_from = mean) |>
    column_to_rownames(var = "gene") |>
    round(digits = 4)
  
  # save avg CPM per group2
  write.table(x = avgPerGroup, 
              file = paste0(out,"counts/avgCPM_per_group_before_filtering.tsv"),
              quote = FALSE,
              sep = "\t")
  
  # all samples in a group2 (i.e. S.8h) must have CPM >= 1
  # create df with summary of if group2 has all samples >= 1  
  filterTable <- cpm |>
    mutate(across(everything(), ~ . >= 1)) |>
    mutate(gene = rownames(cpm)) |>
    pivot_longer(-gene, names_to = "sample_id") |>
    left_join(meta, by = "sample_id") |>
    summarize(thresh = all(value), .by = c(gene, group2)) |>
    pivot_wider(names_from = group2, values_from = thresh) |>
    column_to_rownames(var = "gene") 
  
  # add a 'pass' column that checks if any group2 meets the threshold
  filterTable$pass <- rowSums(filterTable) > 0
  
  # compare row names with 'dge'
  all.equal(rownames(filterTable),rownames(dge))
  
  # filter
  dge.filtered <- dge[filterTable$pass,,keep.lib.sizes = FALSE]
  
  # save
  avgPerGroupFiltered <- avgPerGroup[filterTable$pass,]
  write.table(x = avgPerGroupFiltered, 
              file = paste0(out, "counts/avgCPM_per_group_after_filtering.tsv"),
              quote = FALSE,
              sep = "\t")
  
  # see number of genes leftover
  end <- dim(dge.filtered)[1]
  
  # print message
  print(paste(start - end, "genes removed and", end, "genes leftover."))
  
  # cleanup
  remove(meta,avgPerGroup,cpm,start,end,filterTable,avgPerGroupFiltered)
}
```

## TMM
Convert observed library sizes into effective library sizes. To obtain effective
library sizes, the current library sizes are multiplied by a normalization factor
which we find using the trimmed mean of M values (TMM) method. The TMM method
trims observations from each tail of the distribution of log-rations (M-values). 
By default this method uses a sumTrim of 0.05 and a logratioTrim of 0.3.
```{r TMM_normalize}
# Normalize by trimmed mean of M-values
dge.filtered.norm <- calcNormFactors(dge.filtered, method = "TMM")

# Normalization factor summary
summary(dge.filtered.norm$samples$norm.factors)
```

# QC: After normalization
## Density plot
Density plots of log-intensity distribution of each library can be superposed 
on a single graph for a better comparison between libraries and for 
identification of libraries with weird distribution. 
```{r density_plots}
# set graphical parameter
par(mfrow = c(1,3))

# Normalize data for library size and expression intesntiy
log2cpm.raw <- cpm(dge, log = TRUE)
colnames(log2cpm.raw) <- dge$samples$sample_id
log2cpm.filtered <- cpm(dge.filtered, log = TRUE)
colnames(log2cpm.filtered) <- dge.filtered$samples$sample_id
log2cpm.norm <- cpm(dge.filtered.norm, log = TRUE)
colnames(log2cpm.norm) <- dge.filtered.norm$samples$sample_id

# set colors
colors <- group_colors[dge$samples$group]
nsamples <- ncol(dge)

# First, plot the first column of the log2cpm.raw density
plot(density(log2cpm.raw[,1]), col = colors[1], lwd = 2, ylim = c(0,0.2), 
     las = 2, main = "A. Raw", xlab = expression('Log'[2]~CPM))

# For each sample plot the lcpm density
for (i in 2:nsamples){
  den <- density(log2cpm.raw[,i]) #subset each column
  lines(den$x, den$y, col = colors[i], lwd = 2) 
}

# Second, plot log2cpm.filtered
plot(density(log2cpm.filtered[,1]), col = colors[1], lwd = 2, ylim = c(0,0.25), 
     las = 2, main = "B. Filtered", xlab = expression('Log'[2]~CPM))
abline(v = cpm(3, log = TRUE), lty = 3)
for (i in 2:nsamples) {
  den <- density(log2cpm.filtered[,i])
  lines(den$x, den$y, col = colors[i], lwd = 2)
}

# Third, plot log2cpm.norm
plot(density(log2cpm.norm[,1]), col = colors[1], lwd = 2, ylim = c(0,0.25), 
     las = 2, main = "C. Normalized", xlab = expression('Log'[2]~CPM))
abline(v = cpm(3, log = TRUE), lty = 3)
for (i in 2:nsamples) {
  den <- density(log2cpm.norm[,i])
  lines(den$x, den$y, col = colors[i], lwd = 2)
}

# save
path <- paste0(out, "postfiltering_QC/density_plots")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
remove(den)
```

## Boxplots
```{r boxplots}
# set parameters
par(mfrow = c(1,3))

# First look at dge.tech
boxplot(log2cpm.raw, 
        main="A. Raw", 
        xlab="", 
        ylab=expression('Counts per gene (Log'[2]~'CPM)'),
        axes=FALSE,
        col = colors
        )
axis(2) # 2 = left 
axis(1, # 1 = below 
     at = 1:nsamples, # points at which tick-marks should be drawn
     labels = colnames(log2cpm.raw),
     las = 2,
     cex.axis = 0.8 # size of axis
     )

# Second, look at dge.filtered
boxplot(log2cpm.filtered, 
        main="B. Filtered", 
        xlab="", 
        ylab=expression('Counts per gene (Log'[2]~'CPM)'),
        axes=FALSE,
        col = colors
        )
axis(2)
axis(1, at=1:nsamples,labels=colnames(log2cpm.filtered),las=2,cex.axis=0.8)

# Third, look at dge.norm
boxplot(log2cpm.norm, 
        main="C. Normalized", 
        xlab="", 
        ylab=expression('Counts per gene (Log'[2]~'CPM)'),
        axes=FALSE,
        col = colors)
axis(2)
axis(1,at=1:nsamples,labels=colnames(log2cpm.norm),las=2,cex.axis=0.8)

# save
path <- paste0(out, "postfiltering_QC/counts_per_gene_boxplot")
saveToPDF(paste0(path, ".pdf"), width = 20, height = 6)
remove(log2cpm.raw, log2cpm.norm, log2cpm.filtered)
```

## CPM, FPKM, TPM
```{r postfiltering_cpm_fpkm_tpm}
# CPM, Counts Per Million
# normalize by sequencing depth
cpm <- cpm(dge.filtered.norm)
path <- paste0(out, "counts/CPM_postfiltering.tsv")
write.table(cpm, path, quote = FALSE, sep = "\t")

# FPKM, Fragments Per Kilobase Million
# normalize by sequencing depth and then length
exonic.length <- dge.filtered.norm$genes$exonic_length
exonic.length.kb <- exonic.length / 1000
fpkm <- cpm / exonic.length.kb
fpkm <- sweep(cpm, 1, exonic.length.kb, FUN = '/') # 1 = row
path <- paste0(out, "counts/FPKM_postfiltering.tsv")
write.table(fpkm, path, quote = FALSE, sep = "\t")

# TPM, Transcripts Per Kilobase Million
# normalize by length and then sequencing depth
counts <- dge.filtered.norm$counts
fpk <- counts / exonic.length.kb
sum.col <- colSums(fpk) / 1000000
tpm <- sweep(fpk, 2, sum.col, FUN = '/') # 2 = column
path <- paste0(out, "counts/TPM_postfiltering.tsv")
write.table(tpm, path, quote = FALSE, sep = "\t")

remove(counts,cpm,fpk,fpkm,tpm)
```

## Top transcripts
```{r}
counts <- as.data.frame(dge.filtered.norm$counts)
df <- as.data.frame(rowSums(counts))
colnames(df) <- "rsum"
df$gene <- rownames(df)
df <- df[order(df$rsum, decreasing = TRUE),]
rownames(df) <- 1:nrow(df)
head(df, 20)

path <- paste0(out, "counts/top_transcripts_counts_postfiltering.tsv")
write.table(df, path, quote = FALSE, sep = "\t")
```

# Save
```{r save_final_object, eval=FALSE}
saveRDS(dge.filtered.norm, 
        paste0("../rObjects/", out2, "_DGEList_filtered_normalized.rds"))
```

```{r session_info}
sessionInfo()
```
