---
title: "Psi1 snRNAseq"
subtitle: "Pass 2 Recluster"
author: "Kennedi Todd"
date: "05/19/2025"
output:
  html_document:
  code_folding: hide
theme: cerulean
toc: true
toc_float: true
editor_options: 
  chunk_output_type: inline
---
  
# Setup
## Set working directory
```{r working_directory}
knitr::opts_knit$set(root.dir = ".")
```

## Load libraries
```{r libraries, message=FALSE, warnings=FALSE}
# load libraries
library(cowplot)        # plot_grid()
library(dotenv)         # load_dot_env()
library(dplyr)          # left_join()
library(ggplot2)        # ggplot()
library(gridExtra)      # grid.arrange()
library(rtracklayer)    # import()
library(scCustomize)    # Merge_Seurat_List()
library(Seurat)         # Read10X_h5()
library(SeuratWrappers) # RunPrestoAll
library(stringr)        # str_match()
library(tidyr)          # pivot_wider()
```

## Variables and functions
```{r set_variables_and_thresholds}
# source thresholds and output paths
source("../../refs/thresholds_and_outs.R")
out <- paste0(out, "pass3/")

# single cell functions
files <- list.files("../../functions", full.names = TRUE)
invisible(lapply(files, source))

# load the environment variables
load_dot_env(file = "../../refs/.env")
project_dir <- Sys.getenv("PROJECT_DIR")
```

## Load data
```{r load_data}
mouse.annotated <- readRDS(
  file = paste0( "../../rObjects/seurat_obj_annotated_", filtering_method, "_pass2.rds")
)
```

## UMAP
```{r annotated_umap}
# set colors
cluster_colors <- c("firebrick1","gold","gray","darkgreen","lightblue","cyan","blue",
                    "pink","violet","purple","black")

# annotated umap
DimPlot(mouse.annotated,
        cols = cluster_colors,
        shuffle = TRUE)
```

# Recluster pass 3
## Subset
```{r subset_seurat}
# noisey barcodes
barcodes <- readRDS("../../rObjects/pass2_barcodes_to_remove.rds")

# create keep column
mouse.annotated$keep <- !colnames(mouse.annotated) %in% barcodes
table(mouse.annotated$keep)

# check
DimPlot(mouse.annotated, shuffle = TRUE, group.by = "keep")

# subset on cell types of interest
mouse <- subset(mouse.annotated, keep == TRUE)

# check
DimPlot(mouse, shuffle = TRUE, group.by = "keep")

# cleanup
remove(mouse.annotated)
```

## Filter genes
```{r filter_genes}
# filter genes
counts <- GetAssayData(object = mouse, layer = "counts")
nonzero <- counts > 0  # produces logical
keep <- Matrix::rowSums(nonzero) >= 10  # sum the true/false
counts.filtered <- counts[keep,]  # keep certain genes

# overwrite sub
mouse <- CreateSeuratObject(counts.filtered, meta.data = mouse@meta.data[,c(1,4:27)])

# print features removed
print(paste0(dim(counts)[1] - dim(counts.filtered)[1], " features removed"))

# cleanup data
remove(counts,counts.filtered,nonzero)
```

## Recluster
```{r recluster}
# split
mouse[["RNA"]] <- split(mouse[["RNA"]], f = mouse$sample_id)

# transform
options(future.globals.maxSize = 8000 * 1024^2)
mouse <- SCTransform(mouse, verbose = FALSE)

# PCA
mouse <- RunPCA(object = mouse, assay = "SCT")
DefaultAssay(mouse) <- "SCT"

# re-join layers
mouse[["RNA"]] <- JoinLayers(mouse[["RNA"]])

# elbow plot
ElbowPlot(mouse,
          ndims = 30,
          reduction = "pca") +
  geom_vline(xintercept = 13, linetype = "dashed", color = "red")

# save
saveRDS(mouse, 
        file = paste0( "../../rObjects/seurat_obj_pca_", filtering_method, "_pass3.rds"),
        compress = FALSE)

# run UMAP
mouse.unannotated <- RunUMAP(mouse,
                             dims = 1:13,
                             min.dist = 0.2,
                             n.neighbors = 50,
                             seed.use = 42,
                             reduction = "pca",
                             n.components = 3)

# Determine the K-nearest neighbor graph
mouse.unannotated <- FindNeighbors(object = mouse.unannotated,
                                   assay = "SCT",
                                   reduction = "pca",
                                   dims = 1:13)

# Determine the clusters for various resolutions
mouse.unannotated <- FindClusters(object = mouse.unannotated,
                                  algorithm = 1,
                                  resolution = 0.3)

# Set default assay and normalize for visualization
DefaultAssay(mouse.unannotated) <- "RNA"
mouse.unannotated <- NormalizeData(mouse.unannotated)

# save
saveRDS(mouse.unannotated, 
        file = paste0( "../../rObjects/seurat_obj_unannotated_", filtering_method, "_pass3.rds"),
        compress = FALSE)
```

# Unannotated
## UMAP
```{r unannotated_umap}
Idents(mouse.unannotated) <- "seurat_clusters"

u1 <- DimPlot(mouse.unannotated, shuffle = TRUE, label = TRUE)
u1

u2 <- DimPlot(mouse.unannotated, shuffle = TRUE, dims = c(2,3))
u2
```

```{r save_unannotated_umap}
pdf(paste0(out, "UMAP/unannotated_umap.pdf"), width = 6, height = 5)
u1
dev.off()
```


## Auto markers
```{r unannotated_auto_find_markers}
# set params
Idents(mouse.unannotated) <- "seurat_clusters"

# find markers
markers <- SeuratWrappers::RunPrestoAll(
  object = mouse.unannotated,
  assay = "RNA",
  slot = "counts",
  only.pos = TRUE
)

# filter based on adjusted p-value
markers <- markers[markers$p_val_adj < 0.01,]

# get top 3 and top 20 per cluster
top2 <- Reduce(rbind,
               by(markers,
                  markers["cluster"],
                  head,
                  n = 2))
top20 <- Reduce(rbind,
               by(markers,
                  markers["cluster"],
                  head,
                  n = 20))

# save
write.table(markers, 
            paste0(out, "markers/unannotated_auto_find_markers_adjpval_0.01.tsv"),
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)
```

## Canonical markers
```{r unannotated_canonical_markers}
# Auto markers
d1 <- DotPlot(mouse.unannotated,
              features = unique(top2$gene),
              cols = "RdYlBu",
              assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
d1

# Canonical
canonical_markers <- c(
  "Aldoc", "Aqp4", "Gja1", "Clu", "Apoe",             # Astrocytes
  "Ptprb", "Cldn5", "Flt1",                           # Endothelial
  "Col1a1", "Col1a2", "Dcn",                          # Fibroblast
  "Aif1", "C1qa", "P2ry12", "Tmem119","Trem2","Mrc1", # Microglia
  "Syt1", "Snap25", "Gad1", "Meg3", "Igfbpl1","Sox4", # Neurons
  "Mog", "Opalin", "Mal", "Mag", "Mbp",               # Oligodendrocytes
  "Pdgfra", "Cspg4", "Olig1", "Olig2", "Tnr")         # OPCs

d2 <- DotPlot(mouse.unannotated,
              features = canonical_markers,
              cols = "RdYlBu",
              assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
d2
```

```{r save_unannotated_canonical_markers}
pdf(paste0(out, "markers/unannotated_auto_top2_markers.pdf"), width = 10, height = 10)
d1
dev.off()

pdf(paste0(out, "markers/unannotated_canonical_markers.pdf"), width = 10, height = 10)
d2
dev.off()
```

## Assign idetities
```{r assign_identities}
new_cluster_names <- c(
  "0" = "Neurons 1",
  "1" = "Neurons 1",
  "2" = "Astrocytes",
  "3" = "Neurons 1",
  "4" = "Neurons 1",
  "5" = "Neurons 1",
  "6" = "Neurons 1",
  "7" = "Neurons 1",
  "8" = "Neurons 2",
  "9" = "Neurons 1",
  "10" = "Oligodendrocytes",
  "11" = "Neurons 1",
  "12" = "OPCs",
  "13" = "Neurons 1",
  "14" = "Neurons 3",
  "15" = "Neurons 1",
  "16" = "Neurons 1",
  "17" = "Fibroblasts/Endothelial",
  "18" = "Neurons 1",
  "19" = "Fibroblasts",
  "20" = "Neurons 1",
  "21" = "Ependyma",
  "22" = "Differentiating OPCs",
  "23" = "Neurons 1",
  "24" = "Neurons 1",
  "25" = "Microglia",
  "26" = "Neurons 1"
)

# set new idents
mouse.annotated <- RenameIdents(mouse.unannotated, new_cluster_names)
mouse.annotated$annotated_clusters <- Idents(mouse.annotated)

# set level
mouse.annotated$annotated_clusters <- factor(mouse.annotated$annotated_clusters,
                                  levels = c("Astrocytes",
                                             "Fibroblasts",
                                             "Fibroblasts/Endothelial",
                                             "Microglia",
                                             "Neurons 1",
                                             "Neurons 2",
                                             "Neurons 3",
                                             "OPCs",
                                             "Differentiating OPCs",
                                             "Oligodendrocytes",
                                             "Ependyma"))
Idents(mouse.annotated) <- "annotated_clusters"

# save
saveRDS(mouse.annotated, 
        file = paste0( "../../rObjects/seurat_obj_annotated_", filtering_method, "_pass2.rds"),
        compress = FALSE)
```

# Annotated
## UMAP
```{r annotated_umap}
cluster_colors <- c("firebrick1","gold","gray","darkgreen","lightblue","cyan","blue",
                    "pink","violet","purple","black")

u1 <- DimPlot(mouse.annotated, 
              group.by = "annotated_clusters", 
              cols = cluster_colors,
              shuffle = TRUE)
u1

u2 <- DimPlot(mouse.annotated, 
              dims = c(2,3),
              cols = cluster_colors,
              group.by = "annotated_clusters", 
              shuffle = TRUE)
u2
```

```{r save_annotated_umap}
pdf(paste0(out, "UMAP/annotated_umap.pdf"), width = 6, height = 4)
u1
dev.off()
```

## Auto markers
```{r annotated_auto_find_markers}
# find markers
markers <- SeuratWrappers::RunPrestoAll(
  object = mouse.annotated,
  assay = "RNA",
  slot = "counts",
  only.pos = TRUE
)

# filter based on adjusted p-value
markers <- markers[markers$p_val_adj < 0.01,]

# get top 3 and top 20 per cluster
top2 <- Reduce(rbind,
               by(markers,
                  markers["cluster"],
                  head,
                  n = 2))
top20 <- Reduce(rbind,
               by(markers,
                  markers["cluster"],
                  head,
                  n = 20))

# save
write.table(markers, 
            paste0(out, "markers/annotated_auto_find_markers_adjpval_0.01.tsv"),
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)
```

## Canonical markers
```{r annotated_canonical_markers}
# Auto markers
d1 <- DotPlot(mouse.annotated,
              features = unique(top2$gene),
              cols = "RdYlBu",
              assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
d1

canonical_markers <- c(
  "Aldoc", "Aqp4", "Gja1", "Clu", "Apoe",             # Astrocytes
  "Col1a1", "Col1a2", "Dcn",                          # Fibroblast
  "Ptprb", "Cldn5", "Flt1",                           # Endothelial
  "Aif1", "C1qa", "P2ry12", "Tmem119","Trem2","Mrc1", # Microglia
  "Syt1", "Snap25","Gad1", "Rbfox3","Synpo2", "Prox1",# Neurons
  "Pdgfra", "Cspg4", "Olig1", "Olig2", "Tnr",         # OPCs
  "Mog", "Opalin", "Mal", "Mag", "Mbp",               # Oligodendrocytes
  "Cfap44", "Spag16")                                 # Ependymal

d2 <- DotPlot(mouse.annotated,
              features = canonical_markers,
              cols = "RdYlBu",
              assay = "RNA") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
d2
```

```{r save_annotated_markers}
pdf(paste0(out, "markers/annotated_auto_top2_markers.pdf"), width = 8, height = 10)
d1
dev.off()

pdf(paste0(out, "markers/annotated_canonical_markers.pdf"), width = 8, height = 10)
d2
dev.off()
```

## Marker heatmap
```{r marker_heatmaps}
# Create DimPlot
u <- DimPlot(mouse.unannotated, shuffle = TRUE, cols = cluster_colors) +
  NoLegend()

marker_genes <- c("Aqp4", "Gja1", "Col1a2", "Flt1", "P2ry12","Snap25", "Synpo2",
                  "Rbfox3","Cspg4", "Mog", "Spag16")

# Create 11 FeaturePlots
fplots <- lapply(marker_genes, function(gene) {
  FeaturePlot(mouse.unannotated,
              reduction = "umap",
              features = gene) +
    scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red")) +
    ggtitle(gene) +
    theme(legend.position = "none")
})

# Combine all plots (1 DimPlot + 11 FeaturePlots = 12 total)
all_plots <- c(list(u), fplots)
layout <- matrix(1:12, nrow = 3, ncol = 4, byrow = TRUE)
```

```{r save_marker_heatmap}
# Save as PDF
path <- paste0(out, "markers/umap_heatmap_canonical_markers.pdf")
pdf(path, width = 14, height = 9)
grid.arrange(grobs = all_plots, layout_matrix = layout)
dev.off()
```

# QC
## Heatmap UMAP
```{r umap_heatmap, warning=FALSE, message=FALSE}
# UMAP percent_mt
f1 <- FeaturePlot(mouse.annotated, 
            reduction = "umap", 
            features = "percent_mt")  + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
f1

# UMAP nCount
f2 <- FeaturePlot(mouse.annotated, 
            reduction = "umap",
            features = "nCount_RNA") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
f2

# UMAP nFeature
f3 <- FeaturePlot(mouse.annotated, 
            reduction = "umap", 
            features = "nFeature_RNA") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
f3

# UMAP percent_ribo
f4 <- FeaturePlot(mouse.annotated, 
            reduction = "umap", 
            features = "percent_ribo") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
f4

# UMAP cell_complexity
f5 <- FeaturePlot(mouse.annotated, 
            reduction = "umap", 
            features = "cell_complexity") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
f5

# UMAP percent_hb
f6 <- FeaturePlot(mouse.annotated, 
            reduction = "umap", 
            features = "percent_hb") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
f6
```

```{r save_umap_heatmap, echo=FALSE, eval=FALSE}
path <- paste0(out, "clustering_QC/umap_heatmap_")
pdf(paste0(path, "percent_mito.pdf"), width = 8, height = 6)
f1
dev.off()

pdf(paste0(path, "nCount.pdf"), width = 8, height = 6)
f2
dev.off()

pdf(paste0(path, "nFeature.pdf"), width = 8, height = 6)
f3
dev.off()

pdf(paste0(path, "percent_ribo.pdf"), width = 8, height = 6)
f4
dev.off()

pdf(paste0(path, "cell_complexity.pdf"), width = 8, height = 6)
f5
dev.off()

pdf(paste0(path, "percent_hb.pdf"), width = 8, height = 6)
f6
dev.off()

remove(f1,f2,f3,f4,f5,f6)
```

## Percent cells per cluster
```{r percent_cells}
# timepoint
b1 <- mouse.annotated@meta.data %>%
  group_by(annotated_clusters, timepoint) %>%
  dplyr::count() %>%
  group_by(annotated_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=annotated_clusters,y=percent, fill=timepoint)) +
  theme_classic() +
  geom_col() +
  ggtitle("Percentage of timepoint per cluster") +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust=1))
b1

# sex
b2 <- mouse.annotated@meta.data %>%
  group_by(annotated_clusters, sex) %>%
  dplyr::count() %>%
  group_by(annotated_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=annotated_clusters,y=percent, fill=sex)) +
  theme_classic() +
  geom_col() +
  ggtitle("Percentage of sex per cluster") +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust=1))
b2

# sample
b3 <- mouse.annotated@meta.data %>%
  group_by(annotated_clusters, sample_id) %>%
  dplyr::count() %>%
  group_by(annotated_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=annotated_clusters,y=percent, fill=sample_id)) +
  theme_classic() +
  geom_col() +
  ggtitle("Percentage of sample per cluster") +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust=1))
b3

# phase
b4 <- mouse.annotated@meta.data %>%
  group_by(annotated_clusters, phase) %>%
  dplyr::count() %>%
  group_by(annotated_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=annotated_clusters,y=percent, fill=phase)) +
  theme_classic() +
  geom_col() +
  ggtitle("Percentage of phase per cluster") +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust=1))
b4

# mito.factor
b5 <- mouse.annotated@meta.data %>%
  group_by(annotated_clusters, mito_factor) %>%
  dplyr::count() %>%
  group_by(annotated_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=annotated_clusters,y=percent, fill=mito_factor)) +
  theme_classic() +
  geom_col() +
  ggtitle("Percentage of mito factor per cluster") +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust=1))
b5

# group
b6 <- mouse.annotated@meta.data %>%
  group_by(annotated_clusters, group) %>%
  dplyr::count() %>%
  group_by(annotated_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=annotated_clusters,y=percent, fill=group)) +
  theme_classic() +
  geom_col() +
  ggtitle("Percentage of group per cluster") +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust=1))
b6

# group2
b7 <- mouse.annotated@meta.data %>%
  group_by(annotated_clusters, group2) %>%
  dplyr::count() %>%
  group_by(annotated_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=annotated_clusters,y=percent, fill=group2)) +
  theme_classic() +
  geom_col() +
  ggtitle("Percentage of group2 per cluster") +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust=1))
b7

# dose
b8 <- mouse.annotated@meta.data %>%
  group_by(annotated_clusters, dose) %>%
  dplyr::count() %>%
  group_by(annotated_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=annotated_clusters,y=percent, fill=dose)) +
  theme_classic() +
  geom_col() +
  ggtitle("Percentage of dose per cluster") +
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5, hjust=1))
b8
```

```{r save_percent_cells,echo=FALSE,eval=FALSE}
# save
path <- paste0(out, "clustering_QC/percent_")
pdf(paste0(path, "timepoint_per_cluster.pdf"), width = 8, height = 6)
b1
dev.off()

# save
pdf(paste0(path, "sex_per_cluster.pdf"), width = 8, height = 6)
b2
dev.off()

# save
pdf(paste0(path, "sample_per_cluster.pdf"), width = 8, height = 8)
b3
dev.off()

# save
pdf(paste0(path, "phase_per_cluster.pdf"), width = 8, height = 6)
b4
dev.off()

# save
pdf(paste0(path, "mito_factor_per_cluster.pdf"), width = 8, height = 6)
b5
dev.off()

# save
pdf(paste0(path, "group_per_cluster.pdf"), width = 8, height = 6)
b6
dev.off()

# save
pdf(paste0(path, "group2_per_cluster.pdf"), width = 8, height = 6)
b7
dev.off()

# save
pdf(paste0(path, "age_dose_cluster.pdf"), width = 8, height = 6)
b8
dev.off()

# cleanup
remove(b1,b2,b3,b4,b5,b6,b7,b8)
```

## Cells per sample
```{r cells_per_sample}
data <- as.data.frame(table(mouse.annotated$sample_id))
colnames(data) <- c("sample_id","frequency")

ncells <- ggplot(data, aes(x = sample_id, y = frequency, fill = sample_id)) + 
  geom_col() +
  theme_classic() +
  geom_text(aes(label = frequency), 
            position=position_dodge(width=0.9), 
            vjust=-0.25) +
  scale_y_continuous(breaks = seq(0,40000, by = 10000), limits = c(0,40000)) +
  ggtitle("Filtered: cells per sample") +
  theme(legend.position =  "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
ncells
```

```{r save_cells_per_sample, echo=FALSE, eval=FALSE}
pdf(paste0(out, "clustering_QC/cells_per_sample.pdf"), width = 18, height = 4)
ncells
dev.off()
```

## Cells per cluster
```{r cells_per_cluster}
data <- as.data.frame(table(mouse.annotated$annotated_clusters))
colnames(data) <- c("cluster","frequency")

ncells <- ggplot(data, aes(x = cluster, y = frequency, fill = cluster)) + 
  geom_col() +
  theme_classic() +
  geom_text(aes(label = frequency), 
            position=position_dodge(width=0.9), 
            vjust=-0.25) +
  scale_y_continuous(breaks = seq(0,160000, by = 20000), limits = c(0,160000)) +
  ggtitle("Cells per Cluster") +
  scale_fill_manual(values = cluster_colors) +
  theme(legend.position =  "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
ncells
```

```{r save_cells_per_cluster, echo=FALSE, eval=FALSE}
pdf(paste0(out, "clustering_QC/cells_per_cluster.pdf"), width = 8, height = 6)
ncells
dev.off()
```

## Cells per cluster per sample
```{r cells_per_cluster_per_sample}
# extract data
data <- data.frame(cluster = mouse.annotated$annotated_clusters,
                   sample = mouse.annotated$sample_id)
data_summary <- data %>%
  group_by(cluster, sample) %>%
  summarise(count = n(), .groups = 'drop') %>%
  ungroup()

# Create the stacked bar plot
ncells <- ggplot(data_summary, aes(x = cluster, y = count, fill = sample)) +
  geom_bar(stat = "identity") +
  labs(y = "Frequency", x = "Cluster", fill = "Sample") +
  theme_minimal() +
  ggtitle("Frequency of Cells in Each Cluster with Sample Distribution") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  ylim(c(0,160000))
ncells

# reformat for table output
data_summary <- data_summary %>% pivot_wider(names_from = cluster, values_from = count)
```

```{r save_cells_per_cluster_per_sample, echo=FALSE, eval=FALSE}
pdf(paste0(out, "clustering_QC/cells_per_cluster_with_sample_distribution.pdf"),
    width = 12, height = 6)
ncells
dev.off()
```

## Cells per cluster per group
```{r cells_per_cluster_per_group}
# extract data
data <- data.frame(cluster = mouse.annotated$annotated_clusters,
                   group = mouse.annotated$group)
data_summary <- data %>%
  group_by(cluster, group) %>%
  summarise(count = n(), .groups = 'drop') %>%
  ungroup()

# Create the stacked bar plot
ncells <- ggplot(data_summary, aes(x = cluster, y = count, fill = group)) +
  geom_bar(stat = "identity") +
  labs(y = "Frequency", x = "Cluster", fill = "Group") +
  theme_minimal() +
  ggtitle("Frequency of Cells in Each Cluster with Group Distribution") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  ylim(c(0,160000))
ncells

# reformat for table output
data_summary <- data_summary %>% pivot_wider(names_from = cluster, values_from = count)
```

```{r save_cells_per_cluster_per_group, echo=FALSE, eval=FALSE}
pdf(paste0(out, "clustering_QC/cells_per_cluster_with_group_distribution.pdf"),
    width = 8, height = 6)
ncells
dev.off()

write.table(data_summary, 
            file = paste0(out, "clustering_QC/cells_per_cluster_with_group_distribution.tsv"), 
            sep = "\t",quote = FALSE, row.names = FALSE)
```

# Subcluster neurons
## Subcluster neurons 1
```{r subcluster_neurons}
# subcluster neurons
Idents(mouse.annotated) <- "annotated_clusters"
mouse.annotated <- FindSubCluster(object = mouse.annotated,
                                  graph.name = "SCT_snn",
                                  cluster = "Neurons 1",
                                  resolution = 0.8,
                                  subcluster.name = "neurons1_subcluster")

# reformat
current_ids <- mouse.annotated$neurons1_subcluster
new_ids <- ifelse(grepl("^Neurons 1", current_ids), current_ids, "other")
new_ids <- gsub("Neurons 1_", "", new_ids)
mouse.annotated$neurons1_subcluster <- new_ids
Idents(mouse.annotated) <- "neurons1_subcluster"

# plot
u1 <- DimPlot(mouse.annotated,
              group.by = "neurons1_subcluster",
              label = TRUE)

# extract colors
plot_data <- ggplot_build(u1)$data[[1]]
cluster_order <- levels(Idents(mouse.annotated))
color_mapping <- unique(plot_data[, c("group", "colour")])
color_mapping$group <- factor(color_mapping$group, levels = cluster_order)
color_mapping <- color_mapping[order(color_mapping$group), ]
color_df <- data.frame(Cluster = color_mapping$group, Color = color_mapping$colour)
subcluster_colors <- color_df$Color
subcluster_colors <- subcluster_colors[-length(subcluster_colors)]
subcluster_colors <- c(subcluster_colors,"black")
```

## UMAP
```{r}
u2 <- DimPlot(mouse.annotated,
              group.by = "neurons1_subcluster",
              cols = subcluster_colors,
              label = TRUE)
u2

mouse.annotated$cluster26 <- mouse.annotated$neurons1_subcluster == "26"
u3 <- DimPlot(mouse.annotated,
              group.by = "cluster26",
              label = TRUE)
u3

mouse.annotated$cluster28 <- mouse.annotated$neurons1_subcluster == "28"
u4 <- DimPlot(mouse.annotated,
              group.by = "cluster28",
              label = TRUE)
u4
```

```{r save_annotated_umap}
pdf(paste0(out, "UMAP/neurons1_subclustered_unannotated_umap.pdf"), width = 7, height = 5)
u2
dev.off()

pdf(paste0(out, "UMAP/neurons1_subcluster_26_umap.pdf"), width = 6, height = 5)
u3
dev.off()

pdf(paste0(out, "UMAP/neurons1_subcluster_28_umap.pdf"), width = 6, height = 5)
u4
dev.off()
```

## Extract barcodes
```{r}
barcodes <- rownames(mouse.annotated@meta.data)[mouse.annotated$neurons1_subcluster %in% c("26","28")]
barcodes <- c(barcodes,
              rownames(mouse.annotated@meta.data)[mouse.annotated$annotated_clusters %in% "Ependyma"])

# plot barcodes to keep
mouse.annotated$remove <- rownames(mouse.annotated@meta.data) %in% barcodes
DimPlot(mouse.annotated, group.by = "remove")

# save
saveRDS(barcodes, "../../rObjects/pass2_barcodes_to_remove.rds")
```

